From big data to deep insight in developmental science

### 2015-07-28 wiley-big-data-devel/big-data.tex
nearly complete rough draft

A brash new kid has moved in down the street, and she has stolen the limelight from the old guard.
She's a data scientist, and she's teaching the old dogs of statistics some new tricks about how to detect, describe, and visualize patterns in complex sets of data.
Like successful newcomers in many fields, data scientists build on some well-established ideas and tools, but combine and apply them in new ways to new problems \cite{Press2013b}.
And, like many upstarts in recent memory, data scientists have ridden the wave created by the massive growth in the availability of information over the internet \cite{Borgmann2015}.
In this paper, I will focus on how the increased availability of and interest in analyzing large, complex data sets has influenced the study of human development.
I will begin by discussing the recent history of ``big data" and how, in principle, it has implications for developmental science.
Then, I will survey a set of case studies that illustrate how large, complex data sets have been and are being used currently.
To close, I will discuss a range of issues that the emergence of ``big data" approaches to developmental science pose for the future.

\subsection{``Big Data" is big}

``Big data'' has become a significant cultural phenomenon \cite{borgman_big_2015}, with frequent feature articles in the popular \cite{lohr_big_2012, Marcus2013} and specialist press \cite{HBR2015, Press2013a}.
While much of the public attention focused on big data concerns the influence of social media \cite{meyer_everything_2014}, impacts on economic development \cite{HBR2015} or on data security breaches \cite{Vinton2015}, the topic has also garnered the attention of government officials \cite{NSFBigData}, and private foundations \cite{ArnoldFdnEvidence, GatesFdn, MooreFdn} with interests in funding scientific research.
In March 2012, the U.S. Obama Administration announced an initiative to spend \$200 M dollars on a ``big data'' research and development (R\&D) initiative \cite{Gianchandani2012, Obama2012} with the expressed intent of applying data science techniques to large collections of digital data in order ``to solve some of the Nation's most pressing challenges..." \cite{Obama2012}.
The press release that accompanied the announcement \cite{Obama2012} highlighted some of the big data initiatives already or soon-to-be underway at the federal level, and these include the U.S. National Science Foundation (NSF), U.S. National Institutes of Health (NIH), the Departments of Energy and Defense and the U.S. Geological Survey (USGS).
The European Union (EU) has also taken public positions \cite{EU2013} about the importance of data to fostering social and economic well-being and has made it a priority to grant more open public access to data sets collected or managed by government entities \cite{EU2015}.
These potential benefits for open access to public data extend to governments in the developing world \cite{Malik2015}, and to state and local governments in the U.S.
For example, Indiana has embarked on an effort to reduce infant mortality using big data \cite{Ravindranath2014}.

Notwithstanding the promises of big data to improving public well-being, governments recognize that privacy and security concerns loom large \cite{Obama2014, Vinton2015} and must be addressed to fully realize the promise of big data science, especially in fields focused on human well-being.
For researchers, the promise of government interest in big data extends beyond potentially new or expanded funding sources \cite{NSFBigData}, such as the NIH's Big Data to Knowledge Initiative \cite{BD2K2015} or the now closed National Children's Study \cite{NCS2015}.
It means potentially new or more widely available or readily accessible data sources in open respositories \cite{databrary.org, dataverse.org, ICSPR, macwhinney_childes_2001} accompanied by more stringent requirements for open data sharing from funders \cite{NIMH2015, NSF2011, GatesFdn} and scientific publication outlets \cite{nosek_promoting_2015}.
At the same time, the collection, storage and open sharing of large data sets about pose significant data management challenges that may require changes in technology use \cite{ProjectJupyter, osf} and scientific practice, including closer collaboration between researchers and librarians and information specialists with expertise in data management, storage and long-term preservation \cite{gordon_researcher-library_2015, lynch_big_2008}.
We will revisit these topics later.

On the other hand, developmental scientists have long recognized the importance of collecting and sharing data sets that are big along at least one dimension -- the number of individuals, number of time points sampled, or the number of measures collected \cite{macwhinney_childes_2001}.
In turn, researchers in developmental science continue to pioneer advanced analytical techniques that unpack and explain developmental phenomena across multiple levels of analysis and across wide-ranging timescales \cite{bakeman_sequential_2011, CDS2014, QuantDev}.
Developmental scientists have also been at the vanguard of efforts to create ways to share research data that preserve privacy and uphold ethical research principles \cite{Adolph2012, AERA2011, AERAVideo2015, Asilomar2014, Databrary2015}.
Accordingly, the developmental research community may be especially well-positioned to turn big data into deep insight.

 \subsection{What is big data?}

Doug Laney \cite{laney01controlling3v} was the first to describe the dimensions of data streams that make them especially ``big'' and which accordingly merit special consideration \cite{ibm_2015}: volume, velocity, and variety.
So, data streams may be large in quantity, change rapidly, or consist of many distinct components.
Of course, these dimensions or general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of a specific research context.
What is high volume data to a a developmental psychologist, an archive of 10 terabytes (TB) of video and flat-file data, for example, is tiny to a physicist working on the Large Hadron Collider (LHC) that generates 30 petabytes in a year \cite{CERN_LHC}.
Similarly, the measure of volume matters.
The Inter-university Consortium for Political and Social Research (ICPSR) \cite{ICSPR}, one of the largest data repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections \cite{}
Yet, until the recent acquisition of video data from the Gates Foundation-funded Methods of Effective Teaching (MET) Project \cite{METProject}, the large number of files in the repository totalled about 10 TB of digital storage.
The aspects of volume most pertinent to developmental science include the number of participants or families and the number of measurement time points.
Similarly, the aspects of data velocity most relevant to developmental researchers are those related to the frequency of measurement.
In developmental research, velocity can span many orders of magnitude, from physiological measurements collected at millisecond time scales to longitudinal research spanning years or decades.
Variety encompasses both the range of measurement types employed across developmental research: biological, behavioral, contextual, and cultural/historical.
To the ``three Vs'', variability, or the rate of change over time, and complexity, the mutual interdependence of individual measures, might also be added.

@@

\subsection{Big data about what?}

    % + Biology
    %     * Genes
    %     * Brain activity, structure
    %     * Growth and health
    % + Behavior
    %     * Perception
    %     * Action
    %     * Cognition
    %     * Language
    %     * Emotion / Personality / Temperament
    %     * Friendships, families, & relationships
    %     * Attitudes 
    % + Educational and Employment aptitude/outcomes
    % + Social/economic/cultural context

\subsection{Why big data in developmental science?}
    % + What develops?
    % + Why does it develop that way?
    %     * Changes in the environment?
    %     * Changes in the person?
    % + More precise estimates of small effects
    %     * Trends within person
    %     * across time...
    %     * Example recent replication of GWAS with 100K, <http://pss.sagepub.com/content/early/2014/10/06/0956797614545132.full
    % + Multi-level influences on development
    %     * Influences at different spatial, temporal scales
    %     * Nesting
    % + Improve outcomes for children
    %     * In general
    %     * Disease/disorder focus

@@

        % * Genes
        % 	- EGADs/NEAD (Jenae Neiderhiser)
        %   - Other twin/genetically informed studies 
        % * Brain activity/structure
        %     - Developmental Connectome Project, [website](http://www.developingconnectome.org/)
        %     - David Edwards, Oxford, Imperial College, Kings College, [Centre for the Developing Brain](http://www.kcl.ac.uk/medicine/research/divisions/imaging/centres/cdb/index.aspx)
        %     - John Richards' normative structural MRI series.
        % * Growth and health
        %     - NICHD Study of Early Child Care and Youth Development (SECCYD)
        %     - [National Children's Study](http://www.niehs.nih.gov/research/programs/children-study/), [^11]
        %         - In need of redesign? [^12], according to 2014 report from NAS [^13], 40 academic research centers have been dropped from the program [^14]
        %         - Similar to Japan Environment and Children's Study [^15]
        %         - WHO coordination efforts for Birth Cohort Studies [^16]
        %         - British birth cohort studies [^17], site specificially describing UK longitudinal research datasets [^18]
        %     - World Health Organization (WHO) Child Growth Standards, <http://www.who.int/childgrowth/en/>
        %     - WHO Multicentre Growth Reference Study (MGRS), <http://www.who.int/childgrowth/en/>
        %         - "The MGRS collected primary growth data and related information from approximately 8500 children from widely different ethnic backgrounds and cultural settings (Brazil, Ghana, India, Norway, Oman and the USA)."
        %     - Adolescent Health, <http://www.cpc.unc.edu/projects/addhealth>
        %     - Sex, drugs, alcohol and risky behavior
        %         + Rob Turrisi, Eva Lewkowicz, Chuck Geier
        %         + PSU Prevention Center
        %     - Nutrition

@@

   % + Mass media interest in longitudinal perspective
   %      * Up Series, <http://en.wikipedia.org/wiki/Up_Series>
   %      * Boyhood
   %       * Perception
   %          - Teller Acuity Card norms
   %              + <http://www.ncbi.nlm.nih.gov/pubmed/7890497>
   %              + <http://www.ncbi.nlm.nih.gov/pubmed/7890496>
   %          - Auditory testing norms
   %      * Action
   %          - Gesell?
   %          - Bayley Scales of Infant and Toddler Development
   %              + Nancy Bayley bio, <http://www.pearsonclinical.com/education/authors/bayley-nancy.html>
   %      * Cognition
   %          - Wechsler
   %              + <http://www.iupui.edu/~flip/wechsler.html>
   %              + Wechsler Adult Intelligence Scale (WAIS)
   %              + Wechsler Intelligence Scale for Children (WISC-III), 6-16 year-olds
   %              + Wechsler Preschool and Primary Scale of Intelligence (WPPSI-R), 4-6 1/2 years.
   %      * Language
   %          - Human Speechome, <http://www.media.mit.edu/cogmac/projects/hsp.html>
   %          - [CHILDES/TalkBank](http://talkbank.org/)
   %          - Smith, Aslin, Yu, Frank word learning, grammar
   %          - MacArthur, CDI and Mike Frank's project (wordbank.stanford.edu)
   %          - Cathie Tamis-Lemonda
   %          - Gedeon Deak
   %          - LENA, <http://www.lenafoundation.org/>
   %      * Emotion / Personality / Temperament
   %          - Ecological Momentary Assessment
   %              + Matthias Mehl
   %              + Diary studies
   %          - Infant Behavior Questionnaire, CBC

>>>>

A search on the term ``big data'' yields more than 49 million hits on Google and 13 million hits on Bing.
The results return in less than a second.
The term yields more than 100,000 results on Google Scholar.
A search of the term using Google's ngram viewer (http://books.google.com/ngram) shows the first appearance of the term right before 1900 with a steady rise in frequency from the 1950s to just after 2000.
Google's ngram viewer indexes terms in digitized books.
So, whether measured by search engine matches of electronic documents available on the internet, scholarly documents, or digitized books, the term ``big data'' gets widespread use now.
That use has grown substantially over the past several decades.
The fact that these basic facts about a particular phrase can be determined in an instant speaks to the rapid pace of change in networked computing, search engines, and databases that has occurred largely in the last 20 years (https://en.wikipedia.org/wiki/List_of_websites_founded_before_1995).
``Big data'' has become a significant cultural phenomenon \cite{borgman_big_2015; boyd_critical_2012}, with frequent feature articles in the popular \cite{lohr_big_2012, Marcus2013} and specialist press \cite{HBR2015, Press2013a}.

In this paper, I focus on how the increased availability of and interest in analyzing large, complex, multi-level, multi-measure data sets has influenced, continues to influence, and promises to alter the study of human development.
I begin by asking what makes data especially big, and what implications the size, density, or complexity of datasets might have for understanding human development.
Then, I will briefly survey a set of case studies that illustrate how large, complex data sets have been and are being used currently to reveal the complexities of developmental processes.
To close, I will discuss a range of issues that the emergence of ``big data" approaches to developmental science pose for the future of the field.

I will argue against the claim that the emergence of an era of ``big data'' signals the end of theory \cite{anderson_end_2008} while acknowledging the lure of that notion to some.
Nor, do I find that ``big data'' analyses are all that new in developmental psychology.
Rather, I find that a while the promises of ``big data'' to revolutionize scientific understanding are exaggerated \cite{boyd_critical_2012}, developmental science tackles questions that can benefit substantially from larger, richer, more widely shared, and more readily interrelatable datasets.
Thus, whether significant novel insights into human development emerge from the era of ``big data'' depend not just on the size or complexity of the datasets, but on how widely and openly the data are shared, and how readily able researchers are able to combine or link datasets across levels of analysis.
In turn, these specific innovations depend largely on small, probably manageable, but nonetheless thorny problems related to policy, scientific culture, individual esearcher behavior, publisher priorities, and research funding levels.
Technology may have enabled the ``big data'' era, but the challenges it poses may turn out to be the least important, at least where advancing research in developmental psychology is concerned.

 \subsection{What Does `Big Data` Mean in Developmental Science?}

Doug Laney \cite{laney01controlling3v} was the first person to describe the dimensions of data streams that make them especially ``big'': Volume, velocity, and variety.
Data streams that are large in quantity, change rapidly, or consist of many distinct components merit special consideration \cite{ibm_2015}.
Of course, general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of a specific research context.

What is high volume data to a a developmental psychologist, an archive of 10 terabytes (TB) of video and flat-file data, for example, is tiny to a physicist working on the Large Hadron Collider (LHC) that generates 30 petabytes in a year \cite{CERN_LHC}.
Similarly, the measure of volume matters.
The Inter-university Consortium for Political and Social Research (ICPSR) \cite{ICSPR}, one of the largest data repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections \cite{ICPSR}.
Yet, until the recent acquisition of video data from the Gates Foundation-funded Methods of Effective Teaching (MET) Project \cite{METProject}, the large number of files in the repository totalled about 10 TB of digital storage.
Even more important than the quantity of information stored is the kind of information: What are the targets of inquiry?

Most developmental researchers study individuals or groups of individuals and families.
But, others focus on neighborhoods or schools, or even broader social, cultural, political and environmental influences.
This diversity in the targets of analysis reflects an implicit shared belief that developmental trajectories arise from a variety of influences organized at multiple levels of analysis and operating across multiple time scales.
Consequently, the aspects of `volume' most pertinent to developmental science include the number of participants or families or the number measurement time points.
We will see that many developmental studies involve tens of participants, but some notable datasets have hundreds or thousands of participants.
Similarly, the aspects of data velocity most relevant to developmental researchers are those related to the frequency or spacing of measurements.
In developmental research, velocity can span many orders of magnitude, from physiological measurements collected at millisecond time scales to longitudinal research spanning years or decades.
Variety encompasses both the range of measurement types employed across developmental research: biological, behavioral, contextual, and cultural/historical.
To the ``three Vs'', variability, or the rate of change over time, and complexity, the mutual interdependence of individual measures, might also be added.
For example, patterns of brain activity change rapidly, while working memory spans typically do not.
Brain activity and peripheral autonomic nervous system measurements like heart rate, heart period variability, and galvanic skin responses are related to central nervous system activity in causal, but indirect and complex ways.

@@

\subsection{Big data about what and for what purposes?}

Armed with some definitions about what constitutes big data, we can next ask what topics large datasets can describe and help inform.
The answer spans the range of phenomena covered by developmental science, from genes to geography and from microseconds to millenia.
In the case studies to follow, we will highlight studies that focus on biological evidence from genes, brain structure and activity patterns, hormonal variation, nutrition, growth, and health.
Biological evidence may derive from typically developing or atypically developing populations or both.
% Add citations to examples %
Often, behavioral evidence accompanies biological evidence, sometimes featuring standardized task assessments.
But, there are many examples of big datasets consisting solely of behavioral data.
These provide evidence about patterns of change and stability across the range of phenomena that interest developmental researchers: perception, action, cognition, language, emotion, sleep, personality, temperament, friendship, family, play, school achievement, substance use, sexuality, and sleep.
Still other large datasets focus on educational and employment aptitudes and outcomes and the social, economic, political, cultural, and historical contexts in which children and youth come of age.
% Again find examples and cite %

Large datasets can demonstrate either human diversity or human universals, depending on the topic and area of focus.
In most cases, however, big datasets in developmental science serve to answer core scientific questions that drive research in this area: What develops? 
Why does it develop that way? 
Does the environment change or the person or both, and how so?
Big datasets enable more precise estimation of small effects, for qualities or conditions rare among individuals.
High volume or velocity data may inform the estimation of trends within or between people across time \cite{rietveld_replicability_2014}.
Developmental scientists have long recognized the multiple, nested influences on developmental processes arising at different temporal and spatial scales \cite{elman_rethinking_1998, gottlieb_normally_1998, oyama_ontogeny_2000, vygotsky_mind_1980}.
Only recently have the tools and resources needed to collect high volume, velocity, and variety needed to make many of these theoretical notions empirically testable.
Finally, many, perhaps most big datasets in developmental science have as a primary aim the understanding or treatment of disease or disorder.
In short, big data in developmental science focuses on answering the biggest, most important questions in the field.
In the next section, we turn to evaluating specific cases of research projects that have generated big datasets and data repositories that preserve, store, and share them with the scientific community.

@@

\subsubsection{Epidemiological and Population-based Studies}

Some of the largest and most widely used datasets in developmental science are population-based, often with an epidemiological dimension.
Several notable examples began data collection in the 1950's or 60's.
The National Health and Nutrition Examination Survey (NHANES) \cite{NHANES} began collecting data in 1959 with a focus on characterizing the health and nutritional status of U.S. infants, children, and adults. 
The annual survey combines interviews about demographic, socioeconomic, dietary and health related topics and physical examinations involving medical, dental, and physiological measurements, as well as laboratory tests.
The study assesses a nationally representative sample of about 5,000 persons located across the country, but 15 counties are surveyed annually.
Many will be familiar with some of the study's products such as standardized child growth charts (http://www.cdc.gov/growthcharts/cdc_charts.htm), or related initiatives such as the National Youth Fitness Survey (NNYFS) (http://www.cdc.gov/nchs/nnyfs.htm which collected data on physical activity and fitness levels in order to provide an evaluation of the health and fitness of children in the U.S. ages 3 to 15. 
The NNYFS collected data on physical activity and fitness levels of our youth through interviews and fitness tests.
Much of the NHANES and NNYFS data is available to the public through the National Center for Health Statistics at the U.S. Centers for Disease Control.
However, access to some data are restricted, available only to individuals via a Research Data Center (RDC) that can provide access.
<http://wwwn.cdc.gov/nchs/nhanes/search/variablelist.aspx?CycleBeginYear=1999&Component=Non-Public>.
There are 19 Federal Statistical Research Data Center (RDC) locations across the U.S.
A Google Scholar search shows more than 110,000 publications mention the NHANES dataset.
The Panel Study on Income Dynamics (PSID) \cite{PSID} is another large-scale population based effort.
PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
The study consists of survey responses to questions about employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, education, and other topics. As of mid-2015, PSID had generated more than 3,900 publications \cite{PSID_pubs}. 
Much of data is available publicly to users who register (http://simba.isr.umich.edu/U/ca.aspx) subject to the provisions of a formal set of use conditions (http://simba.isr.umich.edu/U/CondUse.aspx) that focus largely on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data. 
In addition, some restricted data elements are only available subject to a formal restricted use contract (http://simba.isr.umich.edu/restricted/RestrictedUse.aspx) between the investigator and the data owner, the University of Michigan.

In contrast the National Longitudinal Survey of Youth (NLSY) \cite{NLSY} has studied a large subset of the same cohort since the study began in 1979.
The project follows the lives of a sample of almost 10,000 American youth born between 1957-64.
Hosted by the U.S. National Bureau of Labor Statistics (BLS), NLSY ata are now available for up to 25 survey rounds on questions relating to employment and related questions.
The NLSY has also been extended to encompass women born in 1922-37 and 1943-55 and men born 1906-21 and 1941-52 through the National Longitudinal Survey (NLS).
Most NLSY and NLS data are available free of charge to by means of a web-based data portal \cite{NLSY_data}, access to which users must register for.
Users must apply for special permission to access geographic-related variables.
The Center for Human Resource Research (CHRR) at The Ohio State University and the National Opinion Research Center (NORC) at the University of Chicago administer both the data collection and repository functions.
A Google Scholar search in mid-2015 shows more than 20,000 publications cite the NLSY.

The Early Childhood Longitudinal Study (ECLS) is another large-scale population-based study that has followed the same groups of young people over time focusing on child development, school readiness, and early school experiences. 
The ECLS consists of three cohorts, one focused on a sample of children measured beginning at birth and two cohorts following children from kindergarten.
The birth cohort a sample of children born in 2001 and followed from birth through kindergarten entry, and unusual among comparable studies includes data from fathers. 
One of the kindergarten cohorts studied approximately 22,000 children who entered school in 1998-99 and were followe through the eighth grade. 
A second cohort entered kindergarten in 2010-11 and is being followed through the fifth grade.
The project has received funding from the National Center for Educational Statistics (https://nces.ed.gov/ecls/), and the data are archived at ICPSR.
ECLS consists of both public and restricted data, and according to a Google Scholar search has generated more than 8,000 publications, but ICSPR says 167.

The National Longitudinal Study of Adolescent Health (Ad Health) \cite{adolescent_health} was a survey designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors. 
Undertaken in response to a Congressional mandate, the study was carried out in two phases, the first involving surveys of 90,000 7-12th graders at 145 schools around the U.S. during the 1994-1995 school year. 
The second phase involved 20,000 participants, where both the target adolescents and their parents underwent interviews at two time points separated by a year. 
The Ad Health cohort has been followed into young adulthood with four in-home interviews, the most recent in 2008, when the sample was aged 24-32. 
In addition to survey data Add Health includes biomarker data from a blood test taken during Wave IV. 
Data are available about glucose, hemoglobin (HbA1c) levels, and blood lipids. 
The Add Health data \cite{harris_add_health} are archived at ICSPR \cite{ICPSR} and are available for public use.
As of mid-2015, more than 5,000 publications \cite{add_health_publications} had arisen from the Add Health dataset.

In comparison, the NICHD Study of Early Child Care and Youth Development (SECCYD) prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
The goal was to learn more about the kind and quality of child care experiences, and the effects of different child care settings on developmental outcomes. 
Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, and anthropometric measures, and demographic information.
The study included multiple respondents, including the child, parents, caregivers, clinicians, and child care center directors.
Unlike the Ad Health study, data in SECCYD are not available to the public, but are restricted.
Access is granted only by application, and approval is granted on a case-by-case basis.
The SECCYD has generated more than 214 publications as of mid-2015 \cite{seccyd_publications}.

Outside the U.S., there are several examples of large-scale longitudinal studies focused on children's growth and health.
Launched in 2011, the Japan Environment and Children's Study <http://www.env.go.jp/en/chemi/hs/jecs/> is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development. 
Pilot studies started in 2008 and the hundred thousandth participating mother was enrolled in March 2014. 
Health outcomes and exposure measurements will continue until the participating children become 13 years old.
The study is in its early stages, and has only produced a handful of publications <http://www.env.go.jp/en/chemi/hs/jecs/publications/index.html>.
Data sharing plans have not been announced.
World Health Organization (WHO) has offered to help coordinate measurement strategies for large-scale birth cohort studies <http://www.who.int/ceh/cohorts/en/> to ensure that data can be harmonized across them.
Several birth cohort studies have been undertaken in the U.K., beginning with the National Survey of Health & Development in 1946 (https://en.wikipedia.org/wiki/National_Survey_of_Health_%26_Development). 
The National Child Development Study (NCDS) is a continuing longitudinal study that seeks to follow the lives of all those living in Great Britain who were born in one particular week in 1958. 
The origins of the NCDS can be found in the Perinatal Mortality Survey (PMS) which was then sponsored by the National Birthday Trust Fund and set up to collect information about the social and obstetric factors associated with stillbirth and death in early infancy.[3] Examples of other topics which have since been included are medical care, health, home environment, educational progress, parental involvement, family relationships, economic activity, income, training and housing.[4]
Conducted by the Centre for Longitudinal Studies (CLS), the aim of the study is to improve understanding of the factors affecting human development over the whole lifespan. 
It collects information on physical and educational development, economic circumstances, employment, family life, health behaviour, wellbeing, social participation and attitudes.
physical, educational and social development.[1] During the period 2002-2004, genetic information on participants was also obtained to examine the genetic effects on common traits and diseases.[2]
The British Cohort Studies have generated more than 3,500 publications (http://www.cls.ioe.ac.uk/Bibliography.aspx?sitesectionid=647&sitesectiontitle=Bibliography) to-date.
Data are available from (http://ukdataservice.ac.uk/get-data/key-data/cohort-and-longitudinal-studies) by registration.
Another widely known and used dataset comes from the World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS) (http://www.who.int/childgrowth/en/).
The MGRS collected growth and related data from 8500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
The data provide body mass index, height for children 5-19, and weight data norms for children 5-10.

Not all large-scale birth cohort studies succeed.
A notable example in the U.S. is the National Children's Study](http://www.nichd.nih.gov/research/NCS/Pages/default.aspx).
Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children from before birth to age 21. 
However, the NIH Director decided to close the NCS on December 12, 2014 following the recommendations of an advisory panel.
Authorized by the Children’s Health Act of 2000, the NCS was a planned large-scale, long-term study of U.S. children and their mothers and fathers designed to study environmental influences on child health and development. 
Questionnaire, physical measures, biospecimens and environmental sample data from up to 5,726 participants in the NCS Vanguard study were collected in 2009-2014 prior to the NCS study's closure.
Those data are slated for release in a data archive sometime in 2015.

Several themes emerge from this sample of population-based studies.
Datasets are large in terms of the volume of participants sampled ranging in the thousands to tens of thousands.
The datasets demonstrate significant variety of measures, including survey-based and bio-marker-based (NHANES, Ad Health, SECCYD) and even video (SECCYD).
The studies show variety geographic composition, as well.
Some studies involve tens of sample occasions per individual, the NLSY and Ad Health studies in particular.
Datasets are housed either in U.S. Government (NHANES, SECCYD) or university-based repositories (PSID, Ad Health, NLSY).
Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center.
Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact on scientific discovery.
% Susan McHale's comment about training. Check the other datasets for this.
% Limits on geographic linkage, other ways data are anomymized to reduce possibility of identifiability.

\subsubsection{Behavior Genetics}

The next area of developmental science we explore is behavior genetics, where large data samples have long been required in order to detect what can often be small effect sizes.
Several large-scale genetics studies that are not specifically developmental deserve mention.
The Genome of the Netherlands Project(http://www.bbmri.nl/en-gb/activities/rainbow-projects/131-genome-of-the-netherlands) is a publicly available dataset from n=250 families, consisting of two parents plus one of their adult children \cite{Genome_Netherlands}.
The Genes, Environment, and Melanoma Study (GEMS) focuses on risk for melanoma in a sample of 3,700 individuals from the North America, Europe, and Australia (http://gemstudy.org/main/index.html).
The Psychiatric Genomics Consortium (PGC) (http://www.med.unc.edu/pgc) contains data from more than 170,000 individuals who have psychiatric diagnoses or are considered at-risk.
Many disorders featured (autism, attention-deficit hyperactivity disorder, and schizophrenia) in the PGC dataset have developmental dimensions.
Genome-wide association study (GWAS) results from the PGC dataset may be viewed in a specialized web browser tool (http://www.broadinstitute.org/mpg/ricopili/).

The U.K.-based Twins Early Development Study (TEDS) \cite{TEDS} (https://en.wikipedia.org/wiki/Twins_Early_Development_Study) contains survey data from 15,000 English and Welsh families who gave birth to twins between 1994 and 1996.
Some 13,000 pairs remain actively involved, providing lab-based behavioral task and survey responses.
DNA samples from more than 5,000 twins pairs are also available.
TEDS has generated more than 350 publications to-date.
Researchers wishing to gain access to the data may do so by submitting a formal data request specifying the aims of the planned research and the variables that needed satisfy the aims. 
Proposals that do not overlap with analyses already being planned or carried out by the TEDS research team or other collaborators are usually approved.

The Twin and Offspring Study in Sweden (TOSS) \cite{TOSS} focuses on genetic and environmental contributions to measures of family relationships and mental health.
TOSS extends the Twin Moms Project to include a sample of twin fathers (N=320 twin pairs and their families).
In addition, TOSS plans to collect another 250 pairs of twin mothers for a target sample of 900 twin parents (n=3,000 individuals) and their families.
While still relatively early in data collection, more than 24 publications reference the TOSS dataset.
Data sharing plans have not yet been announced.

Illustrative U.S. examples include the Colorado Adoption Project (CAP), the Nonshared Environment in Adolescent Development (NEAD) project, and the Early Growth and Development Study (EGADS).
NEAD studied a sample of 720 families, including twins, siblings, and their parents, in a longitudinal design that involved up to 3 data collection waves.
Researchers focused on survey, self-report, and experimenter ratings (from video) of family relationships, adolescent adjustment and competence.
A Google Scholar search shows 73 publications to-date.
The data will be shared by request to the project team.
The team plans to share the data openly later in 2015.

Adoption studies provide a unique window on the mix of genetic and non-genetic factors that influence child development.
Since its inception in 1976, the Colorado Adoption Project (CAP) (http://ibg.colorado.edu/cap/index.html) has enrolled more than 2,400 participants from more than 450 families.
Data include cognitive abilities, temperament measures, demographic variables, 
Additionally parents completed a questionnaire assessment which included measures of temperament
CAP has generated more than 200 published articles (http://ibg.colorado.edu/cap/publishedarticles.html).
Data collection continues.
The study aims to be the definitive genetically informed longitudinal study of lifespan development \cite{rhea_colorado_2013}.
Data from 1976-1989 are archived at Dataverse (https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/00913) and can be accessed by application to Dataverse.
EGADS is a prospective adoption study of (n=561) triads of birth parents, adoptive parents, and adopted children in the U.S. 
The study focuses on genetic and non-genetic factors that influence children’s adjustment from infancy to age 9. 
The data include self- and other-report measures, DNA, and salivary cortisol.
Data are available by request, with plans for more open data sharing in the near future.

\subsubsection{Brain Imaging}

Several large studies have focused on developmental patterns in brain structure or activity.
For practical reasons -- motion artifacts make imaging very young children extremely challenging -- most projects have focused on children and adolescents.
One of the earliest studies of its kind was a combined longitudinal and cross-sectional study of child and adolescent structural brain development led by Jay Giedd at NIMH \cite{giedd_brain_1999}.
The study collected structural brain information from 145 4 to 20-year-olds along with behavioral and clinical data.
More than 100 of the participants were scanned on more than one occasion, with the time between scans approximately two years.
The original paper describing this study has been cited more than 3,300 times, as of mid-2015.
The dataset has generated other highly cited papers \cite{gogtay_dynamic_2004}, but the data appear not to be available for analyses by researchers outside the original investigative team.
The NIH MRI Study of of Normal Brain Development \cite{NIH_Pediatric_MRI} collected multi-modal structural MRI from 554 children from 4 to 18 years of age.
A second cohort of children from birth to 4 years was scanned longitudinally with up to 10 scans. 
Demographic, hormonal, cognitive, affective, and psychiatric data were also collected.
The data are available to qualified researchers by application, and the dataset has generated some highly cited papers since data collection ended in 2006 \cite{group_total_2012, zielinski_network-level_2010}.
John Richards and colleagues have combined data from the NIH MRI Study, data collected in their own labs, and other public sources to create average structural brain templates that can provide more accurate bases for developmental functional neuroimaging studies using EEG and fMRI \cite{Richards_Developmental_MRI_Database, richards_database_2015}.
These data can be openly shared with appropriate citation.
More recently, the Pediatric Imaging, Neurocognition, and Genetics (PING) Project \cite{PING} collected multimodal neuroimaging data, genotypes, neurodevelopmental histories, and information about cognitive and social and emotional function in more than 1,000 participants 3-20 years of age recruited from Baltimore, Boston, Honolulu, Los Angeles, New Haven, New York, Sacramento, and San Diego. 
The study was funded by the Eunice Kennedy Shriver National Institute of Child Health and Human Development.
Data are available to the research community by application.
As of mid-2015, the project had generated more than 20 publications.
The Developmental Connectome Project \cite{developmental_connectome}, led by investigators at King’s College London, Imperial College London, and Oxford University goes against the trend of imaging studies that focus on older children.
This project aims to study the connectome, a map of human brain connectivity, from 20 to 44 weeks post-conception.
Funded by the European Research Council, the imaging data are accompanied by clinical, behavioral and genetic information.
Investigators hope that the data set will provide a basis for studying genetic and environmental risks that could lead to neurodevelopmental disorders such as Autistic Spectrum Disorder or Cerebral Palsy.
The project has already generated 17 publications as of mid-2015.

Looking across the set of large-scale pediatric brain imaging studies described here, several themes emerge.
One is that these studies require large, complex research teams, usually spanning multiple sites.
Another is that, with some notable exceptions \cite{richards_database_2015}, when the data are available for use by others, that use is controlled by the original investigative team.
In some cases, the original investigative team must be included as an author on any publication.
Another trend among large-scale developmental imaging studies is that the data are usually housed locally, on project-specific sites and not generally on centralized servers that aggregate data across studies and sources \cite{openfmri.org, neurovault.org}.

@@

In the realm of behavior, there are a number of measures that have become standards whose widespread adoption has resulted in data that are large in volume, velocity or variety.
Davida Teller pioneered empirical techniques for measuring visual acuity in preverbal children, and the use of Teller Acuity Cards has resulted in the publication of age-based norms for visual development in infants to four-year-olds \cite{mayer_monocular_1995} based on a sample of more than 400.
A companion study on children in Brazil \cite{salomao_large_1995} was conducted with an even larger sample.
The Bayley Scales of Infant and Toddler Development \cite{bayley2006bayley} arose from the Nancy Bayley's Berkeley Growth Study \cite{jones_berkeley_1941} that studied the behavior, motor coordination, physical activity, cognitive function, and maternal behavior of 61 healthy newborn infants.
The California First Year Mental Scale \cite{bayley1933california} and The California Infant Scale of Motor Development \cite{bayley1936california} the precursors.
The Bayley is widely used in clinical and epidemiological settings such as the EPICure <http://www.epicure.ac.uk/> because of the norms.
However, data about those norms appears to be under the control of the Pearson Publishing company that publishes the Bayley and licenses its use.
Bayley score data are available from the Carolina Abecedarian Project and the Carolina Approach to Responsive Education (CARE), 1972-1992 project on ICPSR (http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/4091?sortBy=19&amp;q=Bayley&amp;searchSource=revise).
The data are freely available to registered users of ICSPR.
